---
title: "Databases, SQL and dplyr"
author: "Dr. Alexander Fisher"
execute:
  warning: true
format: 
    revealjs:
      smaller: true
---

## Numbers every programmer should know

| Task                                | Timing (ns)       | Timing (μs)       |
|-------------------------------------|-------------------|-------------------|
| L1 cache reference                  | 0.5               | 0.0005            |
| L2 cache reference                  | 7                 | 0.007             |
| Main memory reference               | 100               | 0.1               |
| Random seek SSD                     | 150,000           | 150               |
| Read 1 MB sequentially from memory  | 250,000           | 250               |
| Read 1 MB sequentially from SSD     | 1,000,000         | 1,000             |
| Disk seek                           | 10,000,000        | 10,000            |  
| Read 1 MB sequentially from disk    | 20,000,000        | 20,000            |
| Send packet CA->Netherlands->CA     | 150,000,000       | 150,000           |


```{r setup}
#| message: false
#| warning: false
#| include: false
# options(
#   width=70
# )

# knitr::opts_chunk$set(
#   fig.align = "center", fig.retina = 2, dpi = 150,
#   out.width = "100%",
#   warning = TRUE
# )

library(dplyr)
library(RSQLite)
```

::: {.aside}
From [jboner/latency.txt](https://gist.github.com/jboner/2841832) & [sirupsen/napkin-math](https://github.com/sirupsen/napkin-math) <br/>
Jeff Dean's original [talk](http://static.googleusercontent.com/media/research.google.com/en/us/people/jeff/stanford-295-talk.pdf) 
:::


## Implications for big data

Lets imagine we have a *10 GB* flat data file and that we want to select certain rows based on a particular criteria. This requires a sequential read across the entire data set.


| File Location    | Performance                        | Time        |
|:-----------------|:-----------------------------------|:------------|
| in memory        | $10~GB \times (250~\mu s / 1~MB)$  | 2.5 seconds |
| on disk (SSD)    | $10~GB \times (1~ms / 1~MB)$       | 10 seconds  |
| on disk (HD)     | $10~GB \times (20~ms / 1~MB)$      | 200 seconds |

<br/>

This is just for *reading* sequential data, if we make any modifications (*writing*) or the data is fragmented things are much worse. 


## Blocks

<br/>

#### Cost:
::: {.center}
 Disk << SSD <<< Memory
:::


#### Speed:
::: {.center}
Disk <<< SSD << Memory
:::

<br/>

::: {.fragment}
So usually possible to grow our disk storage to accommodate our data. However, memory is usually the limiting resource, and if we can't fit everything into memory?
:::

<p/>

::: {.fragment}
Create *blocks* - group related data (i.e. rows) and read in multiple rows at a time. Optimal size will depend on the task and the properties of the disk.
:::


## Linear vs Binary Search

Even with blocks, any kind of querying / subsetting of rows requires a linear search, which requires $\mathcal{O}(N)$ reads.

. . .

<br/>

We can do better if we are careful about how we structure our data, specifically sorting some (or all) of the columns. 

* Sorting is expensive, $\mathcal{O}(N \log N)$, but it only needs to be done once. 

* After sorting, we can use a binary search for any subsetting tasks, $\mathcal{O}(\log N)$

* In a databases these "sorted" columns are refered to as *indexes*.

* Indexes require additional storage, but usually small enough to be kept in memory even if blocks need to stay on disk.


# Databases

## R & databases - the DBI package

Low level package for interfacing R with Database management systems (DBMS) that provides a common interface to achieve the following functionality:

* connect/disconnect from DB
* create and execute statements in the DB
* extract results/output from statements
* error/exception handling
* information (meta-data) from database objects
* transaction management (optional)


::: {.aside}
See [r-dbi.org](https://www.r-dbi.org/) for more details
:::


## DBI Backends

DBI is a specification, not an implementation, and there are a number of packages that implement the DBI specification for different database systems. [Backends for R-DBI](https://r-dbi.org/backends/) lists all available backends, but some notable ones include:

* `RPostgres`

* `RMariaDB`

* `RSQLite`

* `odbc`

* `bigrquery`

* `duckdb`

* `sparklyr`


## RSQLite

Provides the implementation necessary to use DBI to interface with an SQLite database.


```{r}
library(RSQLite)
```


this package also loads the necessary DBI functions as well (via re-exporting).

. . .

Once loaded we can create a connection to our database,


```{r}
con = dbConnect(RSQLite::SQLite(), ":memory:")
# :memory: creates a temporary, in-memory SQLite data base
str(con)
```



## Example Table
 

```{r}
employees = tibble(
  name   = c("Alice","Bob","Carol","Dave","Eve","Frank"),
  email  = c("alice@company.com", "bob@company.com",
             "carol@company.com", "dave@company.com",
             "eve@company.com",   "frank@comany.com"),
  salary = c(52000, 40000, 30000, 33000, 44000, 37000),
  dept   = c("Accounting", "Accounting","Sales",
             "Accounting","Sales","Sales"),
)
```


. . .


```{r}
dbListTables(con)
```



```{r}
dbWriteTable(con, name = "employees", value = employees)
dbListTables(con)
```



## Removing Tables


```{r}
dbWriteTable(con, "employs", employees)
dbListTables(con)
```


<p/>

. . .


```{r}
dbRemoveTable(con,"employs")
dbListTables(con)
```



## Querying Tables

Databases queries are transactional (see [ACID](https://en.wikipedia.org/wiki/ACID)) and are broken up into 3 steps:


```{r}
(res = dbSendQuery(con, "SELECT * FROM employees"))
```


. . .


```{r}
dbFetch(res)
```


. . .


```{r}
dbClearResult(res)
```



## For convenience

There is also `dbGetQuery()` which combines all three steps,


```{r}
(res = dbGetQuery(con, "SELECT * FROM employees"))
```


## Creating tables

`dbCreateTable()` will create a new table with a schema based on an existing data.frame / tibble, but it does not populate that table with data.


```{r}
dbCreateTable(con, "iris", iris)
(res = dbGetQuery(con, "select * from iris"))
```



## Adding to tables

Data can be added to an existing table via `dbAppendTable()`.


```{r}
dbAppendTable(con, name = "iris", value = iris)
```


. . .


```{r}
dbGetQuery(con, "select * from iris") |> 
  as_tibble()
```


## Recap 

We've connected to a database. We can view the tables within the database: 
```{r}
dbListTables(con)
```

We can even save the connnection to a specific table within the databse:

```{r}
emp <- tbl(con, "employees")
```



## filter() using WHERE

We can run SQL directly:


```{r}
result <- 
  dbGetQuery(con, "SELECT * FROM employees WHERE salary < 40000")

result
```


## group_by() and summarize() via GROUP BY

We can create groups for the purpose of summarizing using `GROUP BY`.

```{r}
result <- 
  dbGetQuery(con, "SELECT dept, COUNT(*) AS n FROM employees GROUP BY dept")

result
```

The "AS" keyword in SQL is essentially a way to rename a column or table in the result. From a dplyr standpoint, it's very similar to how we use `mutate()` or `summarize()`.

## head() using LIMIT 

```{r}
dbGetQuery(con, "SELECT * FROM employees")
```


## Exercise 

Using SQL syntax, 

::: panel-tabset
## Exercise
1. compute the total costs in payroll for this company.

2. the average salary within each department

## Solution

```{r}
total_salary <- dbGetQuery(con, 
"SELECT SUM(salary) AS total_salary
    FROM employees")
total_salary

avg_salary_by_dept <- dbGetQuery(con, 
"SELECT dept,
           AVG(salary) AS avg_salary,
           SUM(salary) AS total_salary
    FROM employees
    GROUP BY dept")
avg_salary_by_dept
```

:::



## Closing the connection


```{r}
con
dbDisconnect(con)
```


<p/>


```{r}
con
```



# dplyr & databases


## Creating a database 

```{r include=FALSE}
unlink("flights.sqlite")
```


```{r}
db = DBI::dbConnect(RSQLite::SQLite(), "flights.sqlite")
( flight_tbl = dplyr::copy_to(
    db, nycflights13::flights, name = "flights", temporary = FALSE) )
```



## What have we created?

All of this data now lives in the database on the *filesystem* not in *memory*,


```{r}
pryr::object_size(db)
pryr::object_size(flight_tbl)
```

```{r}
pryr::object_size(nycflights13::flights)
```


## File size


```{r}
fs::dir_info(glob = "*.sqlite") |>
  select(path, type, size)
```



## What is `flight_tbl`?


```{r}
class(nycflights13::flights)
class(flight_tbl)
```


. . .

::: {.small}
```{r}
str(flight_tbl)
```
:::


## Accessing existing tables


```{r}
dplyr::tbl(db, "flights")
```



## Using dplyr with sqlite 

:::: {.columns .small}
::: {.column width='50%'}
```{r}
(oct_21 = flight_tbl |>
   filter(month == 10, day == 21) |>
   select(origin, dest, tailnum)
)
```
:::
::: {.column width='50%' .fragment}
```{r}
dplyr::collect(oct_21)
```
:::
::::



## Laziness

dplyr / dbplyr uses lazy evaluation as much as possible, particularly when working with non-local backends.

* When building a query, we don't want the entire table, often we want just enough to check if our query is working / makes sense.

* Since we would prefer to run one complex query over many simple queries, laziness allows for verbs to be strung together.

* Therefore, by default `dplyr`

    * won't connect and query the database until absolutely necessary (e.g. show output),

    * and unless explicitly told to, will only query a handful of rows to give a sense of what the result will look like.
    
    * we can force evaluation via `compute()`, `collect()`, or `collapse()`


## dplyr -> SQL - `show_query()`

::: {.small}
```{r}
class(oct_21)
```

```{r}
show_query(oct_21)
```
:::

## More complex queries

:::: {.columns .small}
::: {.column width='50%'}
```{r}
oct_21 |> 
  summarize(
    n=n(), .by = c(origin, dest)
  )
```
:::
::: {.column width='50%' .fragment}
```{r}
oct_21 |> 
  summarize(
    n=n(), .by = c(origin, dest)
  ) |> 
  show_query()
```
:::
::::

##

::: {.small}
```{r}
oct_21 |> 
  count(origin, dest) |> 
  show_query()
```
:::

## SQL Translation 

In general, dplyr / dbplyr knows how to translate basic math, logical, and summary functions from R to SQL. dbplyr has a function, `translate_sql()`, that lets you experiment with how R functions are translated to SQL. 

. . .

::: {.xsmall}
```{r}
#| error: True
con = dbplyr::simulate_dbi()
dbplyr::translate_sql(x == 1 & (y < 2 | z > 3), con=con)
dbplyr::translate_sql(x ^ 2 < 10, con=con)
dbplyr::translate_sql(x %% 2 == 10, con=con)
```
:::

. . .

::: {.xsmall}
```{r}
#| error: True
dbplyr::translate_sql(mean(x), con=con)
dbplyr::translate_sql(mean(x, na.rm=TRUE), con=con)
```
:::

## Complications?

::: {.small}
```{r}
#| error: True
oct_21 |> mutate(tailnum_n_prefix = grepl("^N", tailnum))
```

```{r}
oct_21 |> mutate(tailnum_n_prefix = grepl("^N", tailnum)) |> show_query()
```
:::

## Closing thoughts

The ability of dplyr to translate from R expression to SQL is an incredibly powerful tool making your data processing workflows portable across a wide variety of data backends.

Some tools and ecosystems that are worth learning about:

* Spark - [sparkR](https://spark.apache.org/docs/latest/api/R/index.html), [spark SQL](https://spark.apache.org/docs/latest/api/sql/index.html), [sparklyr](https://spark.rstudio.com/)

* [DuckDB](https://duckdb.org/)

* Apache [Arrow](https://arrow.apache.org/)

